---
title: "Score-based confidence intervals for random effect covariance parameters: FEV1 example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Score-based confidence intervals for random effect covariance parameters: FEV1 example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette illustrates the `reconf` package using a longitudinal lung function
dataset.  We demonstrate score-based confidence intervals for all three covariance
parameters of the random effects: two variances and one covariance.  Standard Wald
intervals do not apply to covariance parameters in general, and profile
likelihood intervals are reported by `lme4` on the Cholesky (SD/correlation)
scale rather than the original variance/covariance scale.  The `reconf` package
provides intervals directly on the natural scale.

## Data

The `fev1` dataset contains 1994 repeated measurements of forced expiratory
volume in one second (FEV1) for $n = 300$ girls and young women from Topeka,
Kansas, originally described in Fitzmaurice, Laird, and Ware (2011).  Each row
is one observation; the variables are:

| Variable   | Description |
|------------|-------------|
| `id`       | Subject identifier |
| `age`      | Age at measurement (years) |
| `ht`       | Height at measurement (metres) |
| `baseage`  | Baseline age (first visit) |
| `baseht`   | Baseline height (first visit) |
| `logfev1`  | Log FEV1 (natural log, litres) |

```{r load-data}
library(reconf)
library(lme4)

data(fev1)
dim(fev1)
head(fev1)
```

Each subject has between 1 and 12 observations over time.

```{r visits-per-subject}
range(table(fev1$id))
```

## Model

We fit the linear mixed model

$$
\text{logfev1}_{it} = \beta_1 + \beta_2\,\text{age}_{it} + \beta_3\,\text{ht}_{it}
  + \beta_4\,\text{baseage}_i + \beta_5\,\text{baseht}_i
  + U_{i1} + U_{i2}\,\text{age}_{it} + E_{it},
$$

where $(U_{i1}, U_{i2})^T \sim N(0, \Psi)$ independently across subjects,
$E_{it} \sim N(0, \psi_4)$ independently of the random effects, and

$$
\Psi = \begin{pmatrix} \psi_{1} & \psi_{2} \\ \psi_{2} & \psi_{3} \end{pmatrix}.
$$

The random effect covariance parameters $(\psi_{1}, \psi_{2}, \psi_{3})$
are of primary interest here.

```{r fit-model}
fit <- lmer(
  logfev1 ~ age + ht + baseage + baseht + (age | id),
  data = fev1,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)
```

## Maximum likelihood estimates

```{r mle}
round(fixef(fit), 3)

as.data.frame(VarCorr(fit), order = "lower.tri")[
  , c("grp", "var1", "var2", "vcov")]
```

The random slope variance $\hat{\psi}_{3}$ is small, making its confidence
interval of particular interest.

## Score-based confidence intervals

`ci_all_lmer()` computes score-based 95% confidence intervals for all
covariance parameters, including the error variance.  Each row is labeled
by the corresponding variance or covariance and shows the estimate
alongside the bounds; the random-effect parameters appear in
lower-triangular order, so `test_idx = 1, 2, 3` refer to $\psi_{1}$,
$\psi_{2}$, $\psi_{3}$.

```{r ci-all, cache = TRUE}
ci <- ci_all_lmer(fit)
ci
```

`ci_lmer()` computes the interval for a single parameter.

```{r ci-single, cache = TRUE}
ci_lmer(fit, test_idx = 3L)   # random slope variance psi_3
```

## Comparison with profile likelihood

For reference, `lme4::confint()` with `method = "profile"` reports profile
likelihood intervals.  Note that lme4 uses a Cholesky parameterization
internally, so its output is on the SD/correlation scale rather than the
variance/covariance scale used by `reconf`.

```{r lr-compare, cache = TRUE}
confint(fit, method = "profile", parm = "theta_")
```

The `.sig01` and `.sig03` rows are standard deviations ($\sqrt{\psi_{1}}$
and $\sqrt{\psi_{3}}$); `.sig02` is the correlation
$\psi_{2}/\sqrt{\psi_{1}\psi_{3}}$.

## Score tests

The intervals above are the package's main purpose; the same machinery also
yields score tests, which we include for completeness.
`score_test_all_lmer()` tests each covariance parameter against zero,
profiling over the remaining parameters.

```{r score-test-all, cache = TRUE}
fit_reml <- lmer(
  logfev1 ~ age + ht + baseage + baseht + (age | id),
  data = fev1, REML = TRUE,
  control = lmerControl(optimizer = "bobyqa")
)
score_test_all_lmer(fit_reml)
```

`score_test_lmer()` tests joint hypotheses at a user-supplied null.  For
example, that the random effects for age contribute nothing:
$H_0: \psi_{2} = \psi_{3} = 0$.  When $\psi_{3} = 0$, the slope $U_{i2}$
is identically zero, which forces $\psi_{2} = 0$ as well; both must be set
to zero in the null.  With REML, the null vector covers only the three
covariance parameters and $\psi_{4}$.

```{r score-test, cache = TRUE}
vc <- as.data.frame(VarCorr(fit_reml), order = "lower.tri")
# Null: psi_2 = psi_3 = 0; psi_1 and psi_4 at estimates
theta_null_reml <- c(vc$vcov[1], 0, 0, sigma(fit_reml)^2)
score_test_lmer(fit_reml, test_idx = c(2L, 3L), theta_null = theta_null_reml)
```

## References

- Fitzmaurice, G. M., Laird, N. M., and Ware, J. H. (2011).
  *Applied Longitudinal Analysis*, 2nd edition. Wiley.
